Validating the perioperative thirst discomfort scale for measuring thirst discomfort prior to procedures

### 2020-08-13 manuscript/index.Rmd
pdf layout with numbers and 1.5 spacing

---
title: "Validating the perioperative thirst discomfort scale for measuring thirst discomfort prior to procedures"
author:
  - Aaron Conway PhD:
      email: aaron.conway@utoronto.ca
      phone: +1 416 946 7112
      address: 585 University Ave | Toronto, ON  M5G 2N2
      institute: [PMCC, LBFON, QUT]
      correspondence: true
  - Kristina Chang MScN:
      institute: [PMCC]
      correspondence: false
  - Megan Bittner BSc:
      institute: [LBFON]
      correspondence: false
  - Dan Phan BSc:
      institute: [LBFON]
      correspondence: false
  - Navpreet Kamboj BScN:
      institute: [LBFON]
      correspondence: false
  - Matteo Parotto MD, PhD:
      institute: [TGH, UoT]
      correspondence: false
  - Amanda Matthews:
      institute: [PMCC]
      correspondence: false
  - Sherryl Alexander:
      institute: [PMCC]
      correspondence: false
  - Pauline Glaves:
      institute: [PMCC]
      correspondence: false
  - Shahvand Masihi:
      institute: [IR]
      correspondence: false
  - Sebastian Mafeld:
      institute: [IR]
      correspondence: false
institute:
  - PMCC: Peter Munk Cardiac Centre, University Health Network, Toronto, Canada
  - LBFON: Lawrence S. Bloomberg Faculty of Nursing, University of Toronto, Toronto, Canada
  - QUT: School of Nursing, Queensland University of Technology (QUT), Brisbane, Australia
  - TGH: Department of Anesthesia and Pain Management, Toronto General Hospital, UHN, Toronto, Canada
  - UoT: Department of Anesthesia and Interdepartmental Division of Critical Care Medicine, University of Toronto, Toronto, Canada
  - IR: Interventional Radiology, JDMI, Toronto General Hospital
bibliography: Extras/references.bib
csl: Extras/sp-basic.csl
output:
  officedown::rdocx_document:
    reference_docx: Extras/manuscript_template.docx
    pandoc_args:
      - --lua-filter=Extras/scholarly-metadata.lua
      - --lua-filter=Extras/author-info-blocks.lua
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)

library(drake)
library(magrittr)
library(flextable)

@@

Symptoms of thirst discomfort are common in patients who are required to undergo long periods of fasting prior to medical and surgical procedures. The aim of this study was to examine the validity and reliability of the perioperative thirst discomfort scale for measuring thirst discomfort prior to procedures.

@@

A prospective observational study was conducted. Patients who were fasted and scheduled for an elective cardiac or interventional radiology procedure were included. Mokken scaling analysis was conducted to investigate the unidimensionality and hierarchical nature of the scale. Correlations between perioperative thirst discomfort scale scores and fasting durations were calculated to evaluate construct validity and convergent validity was evaluated by comparing scores with global thirst discomfort and thirst intensity items.

@@

Responses from 193 participants were included. The mean duration of fasting from food and non-clear fluids was `r food` hours and clear fluids was `r fluids` hours. Five items from the perioperative thirst discomfort scale (PTDS-5) formed a Mokken scale with evidence of invariant item ordering. The median  PTDS-5 score was `r ptds5`  from a possible range of 0 to 10. The hierarchical pattern of responses to PTDS-5 items is interpretable in terms of respondents most easily endorsing an item related to discomfort associated with the desire to drink water through to first endorsing symptoms related to dryness of the mouth and lips before those related to the abnormal sensations of 'thick' saliva and a 'thick' tongue. Scale reliability was adequate with a coefficient score (*rho* =`r drake::readd(reliability_ptds5)`). There was a positive correlation between PTDS-5 scores and the global thirst discomfort rating (*tau* = `r round(global_thirst_corr$tau, 2)`; 95% CI = `r round(global_thirst_corr$CI_low, 2)` to `r round(global_thirst_corr$CI_high, 2)`), as well as thirst intensity (*tau* = `r round(intensity_corr$tau, 2)`; 95% CI = `r round(intensity_corr$CI_low, 2)` to `r round(intensity_corr$CI_high, 2)`). Duration of fasting was not associated with PTDS-5 scores.

@@

The items in the PTDS-5 form a strong Mokken scale, meaning it is a reliable and precise way to order patients according to their thirst discomfort. Nurses may use this simple scale to assess the severity of thirst discomfort in patients who are fasted prior to procedures and initiate interventions according to individual needs.

@@

<div id="refs"></div>

@@

```{r tab.cap="Participant characteristics", tab.id = 'tab1'}
drake::loadd(summary_table)
gtsummary::as_flextable(summary_table)
```

@@

```{r tab.id = "tab2", tab.cap= "Item hierarchy of the PTDS-5"}
flextable::flextable(drake::readd(item_hierarchy_df_ptds5)) %>% 
  autofit()
```

@@

```{r distribution, fig.cap = "Distribution of PTDS-5 scores"}
knitr::include_graphics("figures/distribution.png")
```

@@

```{r intensity, fig.cap="Association between PTDS-5 and thirst intensity. PTDS-5 score ranges from 0 to 10 with higher scores indicating worse thirst discomfort."}
knitr::include_graphics("figures/thirst-intensity.png")
```

@@

```{r globalthirst, fig.cap="Association between PTDS-5 and global thirst rating"}
knitr::include_graphics("figures/global-thirst.png")
```

@@

```{r fasting, fig.cap = "Association between PTDS-5 and duration of fasting"}
knitr::include_graphics("figures/fasting.png")
```

@@

```{r tab.id='Hvaluesptds7', tab.cap.pre="Appendix Table ", tab.cap = "Homogeneity values (and standard errors) for all items in the PTDS"}
drake::readd(H_values_ptds7_table) %>% 
  flextable::flextable() %>% 
  autofit()
```

@@

flextable::flextable(drake::readd(monotonicity_df_ptds6_table)) %>%  
  autofit()
```

@@

```{r tab.id='aisp', tab.cap.pre="Appendix Table ", tab.cap = "aisp for increasing H thresholds (*c*)"}
flextable::flextable(drake::readd(aisp)) %>% 
  autofit()
```

@@

```{r tab.id='iioptds6', tab.cap.pre="Appendix Table ", tab.cap = "Invariant item ordering for PTDS-6"}
drake::readd(iio_ptds6_table) %>% 
  flextable() %>% 
  autofit()
```

@@

```{r tab.id='iioptds5', tab.cap.pre="Appendix Table ", tab.cap = "Invariant item ordering for PTDS-5"}
drake::readd(iio_ptds5_table) %>%
    flextable() %>%
    autofit()
```

>>>>

---
title: "Validating the perioperative thirst discomfort scale for measuring thirst discomfort prior to procedures"
author:
  - Aaron Conway PhD:
      email: aaron.conway@utoronto.ca
      phone: +1 416 946 7112
      address: 585 University Ave | Toronto, ON  M5G 2N2
      institute: [PMCC, LBFON, QUT]
      correspondence: true
  - Kristina Chang MScN:
      institute: [PMCC]
      correspondence: false
  - Megan Bittner BSc:
      institute: [LBFON]
      correspondence: false
  - Dan Phan BSc:
      institute: [LBFON]
      correspondence: false
  - Navpreet Kamboj BScN:
      institute: [LBFON]
      correspondence: false
  - Matteo Parotto MD, PhD:
      institute: [TGH, UoT]
      correspondence: false
  - Amanda Matthews:
      institute: [PMCC]
      correspondence: false
  - Sherryl Alexander:
      institute: [PMCC]
      correspondence: false
  - Shahvand Masihi:
      institute: [IR]
      correspondence: false
  - Sebastian Mafeld:
      institute: [IR]
      correspondence: false
institute:
  - PMCC: Peter Munk Cardiac Centre, University Health Network, Toronto, Canada
  - LBFON: Lawrence S. Bloomberg Faculty of Nursing, University of Toronto, Toronto, Canada
  - QUT: School of Nursing, Queensland University of Technology (QUT), Brisbane, Australia
  - TGH: Department of Anesthesia and Pain Management, Toronto General Hospital, UHN, Toronto, Canada
  - UoT: Department of Anesthesia and Interdepartmental Division of Critical Care Medicine, University of Toronto, Toronto, Canada
  - IR: Interventional Radiology, JDMI, Toronto General Hospital
bibliography: references.bib
csl: Extras/sp-basic.csl
always_allow_html: true

output:
  officedown::rdocx_document:
    reference_docx: Extras/manuscript_template.docx
    pandoc_args:
      - --lua-filter=Extras/scholarly-metadata.lua
      - --lua-filter=Extras/author-info-blocks.lua
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE, fig.retina = 5)
# tint options
library(tint)
knitr::opts_chunk$set(tidy = FALSE, cache.extra = packageVersion('tint'))
options(htmltools.dir.version = FALSE)
library(drake)
library(magrittr)
library(flextable)

@@

Thirst discomfort is common because patients are required to undergo long periods of fasting before medical and surgical procedures. The aim of this study was to examine the validity and reliability of the perioperative thirst discomfort scale (PTDS) for measuring thirst discomfort before procedures.

@@

Fasting patients who were scheduled for an elective cardiac or interventional radiology procedure were included in a prospective observational study. Mokken scaling analysis was used to investigate the dimensionality and hierarchical nature of the PTDS. PTDS score were compated with fasting duration to evaluate construct validity and convergent validity was evaluated by comparing scpre with global thirst discomfort and thirst intensity ratings.

@@

Five items from the perioperative thirst discomfort scale (PTDS-5) formed a Mokken scale with evidence of invariant item ordering. Participants most easily endorsed the item related to the desire to drink water through to first endorsing symptoms related to dryness of the mouth and lips before those related to the abnormal sensations of 'thick' saliva and a 'thick' tongue. Scale reliability was adequate (*rho* =`r drake::readd(reliability_ptds5)`). There was a positive correlation between PTDS-5 scores and the global thirst discomfort rating (*tau* = `r round(global_thirst_corr$tau, 2)`; 95% CI = `r round(global_thirst_corr$CI_low, 2)` to `r round(global_thirst_corr$CI_high, 2)`), as well as thirst intensity (*tau* = `r round(intensity_corr$tau, 2)`; 95% CI = `r round(intensity_corr$CI_low, 2)` to `r round(intensity_corr$CI_high, 2)`). Duration of fasting was not associated with PTDS-5 scores.

@@

The items in the PTDS-5 form a strong Mokken scale, meaning it is a reliable and precise way to order patients according to their thirst discomfort.

@@

::: {#refs}
:::

@@

```{r tab.cap="Participant characteristics", tab.id = 'tab1'}
drake::loadd(summary_table)
# gtsummary::as_flextable(summary_table)
gtsummary::as_kable_extra(summary_table, booktabs = T)
```

@@

```{r tab.id = "tab2", tab.cap= "Item hierarchy of the PTDS-5"}
# flextable::flextable(drake::readd(item_hierarchy_df_ptds5)) %>% 
#   autofit()
drake::readd(item_hierarchy_df_ptds5) %>% 
  kableExtra::kable(booktabs = T)
```

@@

```{r distribution, fig.cap = "Distribution of PTDS-5 scores"}
# knitr::include_graphics("figures/distribution.png")
drake::readd(distribution_plot)
```

@@

```{r intensity, fig.cap="Association between PTDS-5 and thirst intensity. PTDS-5 score ranges from 0 to 10 with higher scores indicating worse thirst discomfort."}
# knitr::include_graphics("figures/thirst-intensity.png")
drake::readd(thirst_intensity_plot)
```

@@

```{r globalthirst, fig.cap="Association between PTDS-5 and global thirst rating"}
# knitr::include_graphics("figures/global-thirst.png")
library(dplyr)
drake::readd(global_thirst_plot)
```

@@

```{r fasting, fig.cap = "Association between PTDS-5 and duration of fasting"}
# knitr::include_graphics("figures/fasting.png")
drake::readd(fasting_duration_ptds_plot)
```

@@

```{r tab.id='Hvaluesptds7', tab.cap.pre="Appendix Table ", tab.cap = "Homogeneity values (and standard errors) for all items in the PTDS"}
drake::readd(H_values_ptds7_table) %>% 
  # flextable::flextable() %>% 
  # autofit()
  kableExtra::kable(booktabs = T)
```

@@

# flextable::flextable(drake::readd(monotonicity_df_ptds6_table)) %>%  
#   autofit()
 drake::readd(monotonicity_df_ptds6_table) %>% 
kableExtra::kable(booktabs = T)
```

@@

```{r tab.id='aisp', tab.cap.pre="Appendix Table ", tab.cap = "aisp for increasing H thresholds (*c*)"}
# flextable::flextable(drake::readd(aisp)) %>% 
  # autofit()
drake::readd(aisp) %>% 
  kableExtra::kable(booktabs = T)
```

@@

```{r tab.id='iioptds6', tab.cap.pre="Appendix Table ", tab.cap = "Invariant item ordering for PTDS-6"}
drake::readd(iio_ptds6_table) %>% 
  # flextable() %>% 
  # autofit()
  kableExtra::kable(booktabs = T)
```

@@

```{r tab.id='iioptds5', tab.cap.pre="Appendix Table ", tab.cap = "Invariant item ordering for PTDS-5"}
drake::readd(iio_ptds5_table) %>%
    # flextable() %>%
    # autofit()
  kableExtra::kable(booktabs = T)
```
